Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-06-16 SIR_manuscript.tex
minor fixes

The expression of $\Rnum$ has a complicated form with all of the model parameters involved. This expression can be simplified and represented given the specific form of matrix $G_{11}$ \eqref{G11}. For the purpose of simplicity we present $\Rnum$ in the manuscript in terms of expressions $C$, $C1$ and $C2$, specified in \eqref{eq:C12}.

@@

{\bf The first goal}
is to explore how changes in isolation, $\theta\_w$ and $\theta\_c$, affects $\Rnum$. Mathematically we would like to verify the sign of $\pder \Rnum{\theta\_w}$ and $\pder \Rnum{\theta\_c}$. We start with simplifying $\Delta$ \eqref{eq:del4} by factoring $\theta\_w$ and $\theta\_c$ in Eq.~\eqref{eq:del4}. Thus, $\Delta$ can be rewritten as
\begin{equation}
\label{eq:del4_theta}
\Delta= \frac{1}{C}\Big(
-D_1 S\_n^* \boldsymbol {\theta\_w^2}+
\big(-\omega^2 p_I \hat F_I S\_n^* \boldsymbol{\theta\_c} + D_2 S\_n^* + \gamma \hat F_I(\omega+\gamma) \big)\boldsymbol{\theta\_w}+
(\omega+\gamma S\_u^*)\omega p_I \hat F_I \boldsymbol{\theta\_c}
\Big),
\end{equation}
where
\begin{align}
\label{eq:del4_d1d2}
D_1=& (\omega+\gamma)\gamma^2+(\omega+\gamma+\omega p_I)\gamma \hat F_I, \\
D_2=& (3\omega+2\gamma)\gamma^2+(\omega+\gamma+2\omega p_I)\gamma \hat F_I+(\gamma+p_I \hat F_I)\omega^2.
\end{align}
$\Delta$, Eq.~\eqref{eq:del4_theta}, is linear in $\theta\_c$ with a positive coefficient. thus
\begin{equation}
\label{eq:d_del4_thetac}
\pder\Delta{\theta\_c}=1/C(\gamma S\_u^*+\omega(1-\theta\_w S\_n^*))\omega p_I \hat F_I.
\end{equation}
This results in increasing $\Delta$, thus decreasing $\Rnum$ with respect to $\theta\_c$, that is $\pder \Rnum{\theta\_c}\leq 0$. Note that $C$ is independent of $\theta\_c$ and $\theta\_w$.

With a similar logic, $\Delta$ \eqref{eq:del4_theta} is a concave-down quadratic equation in $\theta\_w$, given by
\begin{equation}
\label{eq:del4_thetaw}
1/C \Big( -D_1 S\_n^* \boldsymbol {\theta\_w^2}+
\big(-\omega^2 p_I \hat F_I S\_n^* \boldsymbol{\theta\_c} + D_2 S\_n^* + \gamma \hat F_I(\omega+\gamma) \big)\boldsymbol{\theta\_w} \Big).
\end{equation}
We show that the feasible range of $\theta\_w$ lies between 0 and the vertex of this parabola where the parabola is increasing in $\theta\_w$, and so does $\Delta$ which results in inferring $\pder\Rnum{\theta\_w}\leq 0$.
It is enough to show that partial derivative of the expression \eqref{eq:del4_thetaw} with respect to $\theta\_w$ at $\theta\_w=1$ is non-negative. It follows that
\begin{align}
\label{eq:d_del4_thetaw}
\pder\Delta{\theta\_w}\bigg\rvert_{\theta\_w=1} =&
1/C\Big( (D_2-2D_1-\omega^2 p_I \hat F_I \theta\_c) S\_n^*+\gamma \hat F_I(\omega+\gamma) \Big) \\\notag
=& 1/C\Big( (\gamma(\omega+\gamma)+\gamma\omega^2+(1-\theta\_c)\omega^2 p_I \hat F_I ) S\_n^*
+\gamma(\omega+\gamma) \hat F_I (1-S\_n^*) \Big),
\end{align}
which is a positive quantity, given that $\theta\_c$ and $S\_n^*$ vary between 0 and 1.

@@

Having presented the formal analysis, we now concern ourselves with its \textbf{biological interpretations.} We begin by interpreting \ref{case_1_ineq}, under which returning tests more rapidly is \emph{always} harmful. Notice that the ratio $\frac{\theta\_c}{\theta\_w}$ is simply a measure of how much more strongly individuals self-isolate when they test positive compared to when waiting for tests. Since the rate of test return directly influences the rate at which individuals change from a waiting state to a confirmed-positive state, it is intuitive that $\frac{\theta\_c}{\theta\_w}$ would appear in \ref{case_1_ineq}. Next, note that the left-hand side increases when test sensitivity decreases and when targeting of positive individuals is poor. This is consistent with our intuition: a false negative that is returned more rapidly may encourage an infectious individual to not carefully self-isolate, thus increasing transmission. Likewise, if individuals tested are mainly susceptible (rather than infectious), then returning tests more slowly would encourage them to self-isolate for longer while awaiting test results. Having understood the role of each of the parameters in \ref{case_1_ineq}, a ``holistic" interpretation of this inequality is that returning tests more slowly is helpful when the benefit of extended self-isolation (due to awaiting test results) outweighs the benefit of identifying positive cases.

>>>>

The expression of $\Rnum$ has a complicated form with all of the model parameters involved. This expression can be simplified and represented given the specific form of matrix $G_{11}$ \eqref{G11}. For the purpose of simplicity we present $\Rnum$ in the manuscript in terms of expressions $C$, $C_1$ and $C_2$, specified in \eqref{eq:C12}.

@@

{\bf The first goal}
is to explore how changes in isolation, $\theta\_w$ and $\theta\_c$, affects $\Rnum$. Mathematically we would like to verify the sign of $\pder \Rnum{\theta\_w}$ and $\pder \Rnum{\theta\_c}$. We start with simplifying $\Delta$ \eqref{eq:del4} by factoring $\theta\_w$ and $\theta\_c$ in Eq.~\eqref{eq:del4}. Thus, $\Delta$ can be rewritten as
\begin{equation}
\label{eq:del4_theta}
\Delta= \frac{1}{C}\Big(
-D_1 S\_n^* \boldsymbol {\theta\_w^2}+
\big(-\omega^2 p_I \hat F_I S\_n^*\theta\_c+ D_2 S\_n^*+\gamma \hat F_I(\omega+\gamma) \big)\theta\_w+
(\omega+\gamma S\_u^*)\omega p_I \hat F_I \theta\_c
\Big),
\end{equation}
where
\begin{align}
\label{eq:del4_d1d2}
D_1=& (\omega+\gamma)\gamma^2+(\omega+\gamma+\omega p_I)\gamma \hat F_I, \\
D_2=& (3\omega+2\gamma)\gamma^2+(\omega+\gamma+2\omega p_I)\gamma \hat F_I+(\gamma+p_I \hat F_I)\omega^2.
\end{align}
$\Delta$, Eq.~\eqref{eq:del4_theta}, is linear in $\theta\_c$ with a positive coefficient. thus
\begin{equation}
\label{eq:d_del4_thetac}
\pder\Delta{\theta\_c}=1/C(\gamma S\_u^*+\omega(1-\theta\_w S\_n^*))\omega p_I \hat F_I.
\end{equation}
This results in increasing $\Delta$, thus decreasing $\Rnum$ with respect to $\theta\_c$, that is $\pder \Rnum{\theta\_c}\leq 0$. Note that $C$ is independent of $\theta\_c$ and $\theta\_w$.

With a similar logic, $\Delta$ \eqref{eq:del4_theta} is a concave-down quadratic equation in $\theta\_w$, given by
\begin{equation}
\label{eq:del4_thetaw}
1/C \Big( -D_1 S\_n^*\theta\_w^2+
\big(-\omega^2 p_I \hat F_I S\_n^*\theta\_c+D_2 S\_n^*+\gamma \hat F_I(\omega+\gamma) \big)\theta\_w\Big).
\end{equation}
We show that the feasible range of $\theta\_w$ lies between 0 and the vertex of this parabola where the parabola is increasing in $\theta\_w$, and so does $\Delta$ which results in inferring $\pder\Rnum{\theta\_w}\leq 0$.
It is enough to show that partial derivative of the expression \eqref{eq:del4_thetaw} with respect to $\theta\_w$ at $\theta\_w=1$ is non-negative. It follows that
\begin{align}
\label{eq:d_del4_thetaw}
\pder\Delta{\theta\_w}\bigg\rvert_{\theta\_w=1} =&
1/C\Big( (D_2-2D_1-\omega^2 p_I \hat F_I \theta\_c) S\_n^*+\gamma \hat F_I(\omega+\gamma) \Big) \\\notag
=& 1/C\Big( (\gamma(\omega+\gamma)+\gamma\omega^2+(1-\theta\_c)\omega^2 p_I \hat F_I ) S\_n^*
+\gamma(\omega+\gamma) \hat F_I (1-S\_n^*) \Big),
\end{align}
which is a positive quantity, given that $\theta\_c$ and $S\_n^*$ vary between 0 and 1.

@@

Having presented the formal analysis, we now concern ourselves with its \textbf{biological interpretations.} We begin by interpreting \ref{case_1_ineq}, under which returning tests more rapidly is \emph{always} harmful. Notice that the ratio $\frac{\theta\_c}{\theta\_w}$ is simply a measure of how much more strongly individuals self-isolate when they test positive compared to when waiting for tests. Since the rate of test return directly influences the rate at which individuals change from a waiting state to a confirmed-positive state, it is intuitive that $\frac{\theta\_c}{\theta\_w}$ would appear in \ref{case_1_ineq}. Next, note that the left-hand side increases when test sensitivity decreases and when targeting of positive individuals is poor. This is consistent with our intuition: a false negative that is returned more rapidly may encourage an infectious individual to not carefully self-isolate, thus increasing transmission. Likewise, if individuals tested are mainly susceptible (rather than infectious), then returning tests more slowly would encourage them to self-isolate for longer while awaiting test results. Having understood the role of each of the parameters in \ref{case_1_ineq}, a holistic interpretation of this inequality is that returning tests more slowly is helpful when the benefit of extended self-isolation (due to awaiting test results) outweighs the benefit of identifying positive cases.
